The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2020-03-31 razzo-discussion.tex
Good Enough

We expected to find a bigger inference error 
when increasing the amount of co-occurring speciation events.
The reason is simple: the tree prior used assumes speciation does not
co-occur. The stronger the amount of co-occurring speciation, the stronger
the deviation from the species tree prior.

\subsection{Research}

@@

\subsection{Methodological choices}

When simulating a DNA alignment, we chose to use the simplest (JC69) site
model and the simplest (strict) clock model. One could argue these models
are overly simplistic and biologically irrelevant. We argue that this
actually improves the clarity of our results for two reasons. The first
reason is a practical one: because the computational load was lowered,
we were able to perform more replicates. For a process with this high
amount of stochasticity, we think this is highly favored. The second reason
is that we think this reduces the noise of our results, without 
hurting the experiment: in our setup, it is essential
to assume the same generative site and clock model as the one that is actually 
used. Sure, we could have generated the alignment with a 
complex (GTR) site model and a complex (relaxed log-normal) clock model, then
correctly assume these were the generative site and clock model.
But that would be besides the point of the experiment: to measure the impact
of the tree prior.

When simulating our phylogenies, we used an MBD model with and without
extinction. In the case in which extinction is present,
picking the BD tree prior as the generative tree prior seems justified:
it is the best-fitting standard tree prior. For the MBD model without
extinction, we could have picked the Yule tree prior. 
We predicted that the Yule tree prior will be selected as the best candidate model
more often in this scenario, which shows the sensitivity of the method we used.

To compare multiple inference models, we picked the Nested Sampling
approach to directly compare marginal likelihoods. We could have used
alternative methods like \richel{forgot the term for now: where you
do an MCMC that can switch between models}. The reason we used nested
sampling was mostly pragmatic and we think other methods are equivalent.

We used the absolute nLTT statistic to measure the difference between
two phylogenies. Because the crown age of all phylogenies are equal,
normalization on the time has no effect at all. Also because of these
same crown ages, we could have picked many tree difference metrics, like
\richel{must check:} Pagel's delta or Pigot's delta-r. The reason we used
the nLTT statistic was mostly pragmatic. \verb;pirouette; does allow to 
use a different tree difference metric.

@@

The crown age we selected for our experiments was based on trial and error
of pilot experiments. Because the number of species in an MBD process is
expected to increase exponentially, increasing the crown age has a
profound effect on the number of taxa, with the consequence of having
runs that took days to calculate. We decided to prefer a lower number of
taxa at the gain of having more replicates. As a consequence, 
this made our research findings 
\richel{that is, I predict, a bigger error in settings with
more co-occurring speciation} more likely to happen: we know that the
choice of a Bayesian prior is most important in weaker data.

The DNA sequence length we used was 1000 base pairs. This is a DNA
sequence length that we \richel{at least I} feel is informative enough
for the amount of taxa we simulate. Additionally (although sequencing
techniques advance at a great pace), this is -to our knowledge- 
of the same magnitude as the DNA length sequenced in the field.

Ideally, our ESSes would all be above the recommended value of 
200 \richel{I predict not all will be above 200, simply due to 
stochasticity}. We are aware ESSes can be increased easily by
making the MCMC chain longer. The shape of the ESS distribution
(see figure \ref{fig:esses_cumulative}), however, is disadvantageous
to do so: to reduce the percentage of $ESS < 200$ by fifty percent,
we would have to increase the MCMC chain lengths by a factor of forty.
We invested this run-time in doing more replicates.

We compare our generative to model to \richel{I predict we will} 
a limited set of candidate models. Also this was for practical
reasons.

@@

Because the MBD tree prior has only recently been 
developed, it's biological relevance is unknown.
Would the MBD tree prior be employed on biological data of a system
in which co-occuring speciation is known to have taken a strong role,
it will be hard to conclude its performance is superior to a
rate-shift model or an early-burst model. One way to do so, is
to add the MBD other speciation models as a species tree prior
in a phylogentic tool and do a model comparision (e.g. using Nested Sampling).
Another way to do so, is measure the impact a rate-shift or early-burst model
has on the inference error in the same way as this research.
Doing so would also allow to put the quantitative results into perspective,
helping us in how to judge the error distributions.

Expressing the inference error of an MBD process as a distribution
is a good first step in judging the impact an MBD process has on
our predictions, when compared to standard models. It is only a first
step as one still needs to subjectively judge these error distributions,
which was exactly what \verb;pirouette; was hoping to avoid. 
One way to improve our judgement, is by testing it on known species tree
priors, for example, by simulating true trees with a BD model, when assuming
a Yule tree prior.

The effect of using the absolute nLTT tree statistic to quantify the error
is unknown. A pilot experiment \richel{pirouette example 14} shows that
using the logarithmically transformed nLTT statistic gives a messier and
flatter error distribution, but this is not to be taken for granted.

We chose to use few taxa (as we preferred more replicates), but we do not
know the results would the data be more informative (that is, have more taxa).
A cool feature of \verb;pirouette; is that it can quantify when this happen;
when the prior is uninformative due to strong data.

>>>>

We expected to find a bigger inference error 
when increasing the amount of co-occurring speciation events.
The reason is simple: the tree prior used assumes speciation does not
co-occur. The stronger the amount of co-occurring speciation, the stronger
the deviation from the species tree prior. That we did not find this, 
is unexpected.

The first reason why this may be the case, is because of the noise
generated by the runs that did not converge. We expect that the bimodality
in the error distribution is caused by most runs converging (at the lower
errors) and runs with low ESSes (at the higher errors). Filtering away
the runs with a low ESS may help to get a better idea if the expected
inference error still cannot be detected.

The second reason why this we did not detect an inference error may
be due to the choice of our error measuring method: by using the nLTT
statistic. Another method may be better at detecting an inference error.

A third reason might be that the MBD parameter settings chosen, resulted
in too few multiple-birth events. We chose low values of $\nu$ to
prevent doing inference on huge trees. Being able to work on bigger
trees might resolve this problem.

@@

For our experiment, we used the default \verb;pirouette; setup:
when simulating a DNA alignment, the simplest (JC69) nucleotide
substitution model and the simplest (strict) clock model are used. 
One could argue these models
are overly simplistic and biologically irrelevant. We argue that this
actually improves the clarity of our results for two reasons. The first
reason is a practical one: because the computational load was lowered,
we were able to perform more replicates. For a process with this high
amount of stochasticity, we think this is highly favored. The second reason
is that we think this reduces the noise of our results, without 
hurting the experiment: in our setup, it is essential
to assume the same generative site and clock model as the one that is actually 
used.

When simulating our phylogenies, we used an MBD model with and without
extinction. In the case in which extinction is present,
picking the BD tree prior as the generative tree prior seems justified:
it is the best-fitting standard tree prior. For the MBD model without
extinction, we could have picked the Yule tree prior. We expect, however,
that this is of minor relevance.

@@

The crown age we selected for our experiments was based on trial and error
of pilot experiments. Because the number of species in an MBD process is
expected to increase exponentially, increasing the crown age has a
profound effect on the number of taxa, with the consequence of having
runs that took days to calculate. We decided to prefer a lower number of
taxa at the gain of having more replicates.

Ideally, our ESSes would all be above the recommended value of 
200. We are aware ESSes can be increased easily by
making the MCMC chain longer. However, the bimodal
distribution of ESSes is disadvantageous: to reduce the percentage 
of $ESS < 200$ by fifty percent,
we would have to increase the MCMC chain lengths by a factor of forty.
We invested this run-time in doing more replicates.

@@

Because the MBD tree prior has only recently been 
developed, it's biological relevance is unknown.
Would the MBD tree prior be employed on biological data of a system
in which co-occuring speciation is known to have taken a strong role,
it will be hard to conclude its performance is superior to a
rate-shift model or an early-burst model. One way to do so, is
to add the MBD other speciation models as a species tree prior
in a phylogentic tool and do a model comparision (e.g. using nested sampling).
Another way to do so, is measure the impact a rate-shift or early-burst model
has on the inference error in the same way as this research.
Doing so would also allow to put the quantitative results into perspective,
helping us in how to judge the error distributions.
